Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-03 paper.tex
Updates from overleaf

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional information. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritance
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. On the surface, it would appear that \msprime\ is also guilty of this
lack of modularity since it supports simulating both ancestry and mutations.
However, this is essentially a historical accident and because of the \tskit\
library, the mutation process in \msprime\ could just as easily
be performed in a separate package.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

@@

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality and correctness. As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor updates the codebase), using standard Continuous
Integration (CI) methodology. Other CI services check code formatting
and for common errors, as well as producing reports on the level of
test coverage, e.g., pinpointing areas where additional unit tests may
be required.

@@

>>>>

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional information. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritance
between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. Although \msprime\ also supports simulating both ancestry and mutations,
the two aspects are functionally independent within the software.
Both ancestry and mutation simulators are present in \msprime
for reasons of convenience and history,
and could be split into separate packages.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

@@

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor updates the codebase), using standard Continuous
Integration (CI) methodology. Other CI services check code formatting
and for common errors, as well as producing reports on the level of
test coverage, e.g., pinpointing areas where additional unit tests may
be required.

@@

More visibly to the end user, we also have a high standard for documentation,
with precise, comprehensive, and cross-linked documentation
that is automatically built from the code base
and served through the website \url{https://tskit.dev}.
With the goal of lowering the entry barrier to new users,
we have invested significant effort in writing examples and introductions,
and making common tasks discoverable.
We also view contributions to documentation as equally important to the project
as writing code or designing methods---what use would it be
to write reliable, stable software if no-one used it?
